How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
Avoid "in sync" and "out of sync".

The observation that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, which adds a
JIT-compiler to CPython.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} for
the CLR.  All of these need to be kept in sync with the relatively fast
evolution of the language.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them in sync.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/POSIX to the JVM to .NET.

@@

This result was one of the major goals and motivations for the whole
approach.  By construction, the JIT does not get out of sync when the
language evolves, and any code written in the dynamic language runs
correctly under the JIT.  Some very simple Python examples run more than
100 times faster.  At the time of this writing this is still rather
experimental, and the techniques involved are well beyond the scope of
the present paper.  The reader is referred to \cite{D08.2} for more
information.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has primarily only
advantages, among them the avoidance of a proliferation of implementations
growing out of sync.  Writing interpreters both flexibly and efficiently
is difficult and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the Java and .NET.

>>>>

The observation that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, which adds a
JIT-compiler to CPython.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} for
the CLR.  All of these need to be kept synchronized with the relatively fast
evolution of the language.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them synchronized.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/POSIX to the JVM to .NET.

@@

This result was one of the major goals and motivations for the whole
approach.  By construction, the JIT stays synchronized with its VM
and with the language when it evolves,
and any code written in the dynamic language runs
correctly under the JIT.  Some very simple Python examples run more than
100 times faster.  At the time of this writing this is still rather
experimental, and the techniques involved are well beyond the scope of
the present paper.  The reader is referred to \cite{D08.2} for more
information.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has primarily only
advantages, among them the avoidance of a proliferation of diverging
implementations.  Writing interpreters both flexibly and efficiently
is difficult and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the Java and .NET.
